Relative parabolic factorizations of Chevalley groups

### 2015-08-20 congrsubgr-boundgen.tex
Added more items to bibliography; added some text to introduction

By Lemma~\ref{lemma:Stepanov-ideal} the group $\E\left(\Phi,\mathcal{O}_S,I^2\right)$ is contained in $\E(\Phi,I)$ for any root system $\Phi\neq\rC_\ell$ of rank $\geqslant2$.
Notice that under the assumptions of the lemma it is known that $\E(\Phi,I)$ has finite width with respect to $\{x_\alpha(s)\mid\alpha\in\Phi,\ s\in I\}$, see~\cite[Theorem~3.3]{TavgenThesis}.

>>>>

By Lemma~\ref{lemma:Stepanov-ideal} the group $\E\left(\Phi,\mathcal{O}_S,I^2\right)$ is contained in $\E(\Phi,I)$ for any root system $\Phi\neq\rC_\ell$ of rank $\geqslant2$.
Notice that under the assumptions of the lemma it is known that $\E(\Phi,I)$ has finite width with respect to $\{x_\alpha(s)\mid\alpha\in\Phi,\ s\in I\}$, see~\cite{Tavgen93} or \cite[Theorem~3.3]{TavgenThesis}.

### 2015-08-20 paper.tex
Gauss for sr(I)=1
Added more items to bibliography; added some text to introduction

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and let $\E(\Phi, R)$ denote its elementary subgroup, 
 i.\,e. the group generated by root unipotents $t_\alpha(\xi)$, $ $ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the relative Chevalley group (or the principal congruence subgroup) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item Gauss decomposition i.\,e. decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item Bass---Kolster decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$ is the maximal parabolic subgroup of $\G$ corresponding to $i$-th vertex of the Dynkin diagram of $\Phi$ and 
      $\U_i$, $\U_i^-$ are the unipotent radicals associated with it;
\item Dennis---Vaserstein decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over rings having small dimension.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, 
 while Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions such as the stable range condition.

Another advantage of parabolic factorizations is that they prove useful in the study of asymptotical properties of Chevalley groups.
The following theorem is the first such application.

Let $G$ be an abstract group with a generating set $X \subset G$. %(for simplicity we assume $X=X^{-1}$ and $1\in X$),
One defines {\it the width of $G$ with respect to $X$} as the smallest natural number $N=w(G,X)$ such that every element of $G$ is a product of at most $N$ elements of $X$.

The relative elementary group $\E(\Phi,R,I)$ is generated by certain elements $z_\alpha(s,\xi)$, see Theorem~\ref{theorem:Tits-Vaserstein}. The width $w(\E(\Phi,R),\mathcal{X})$, where $\mathcal{X}$ is the set of all elementary generators $x_\alpha(\xi)$, has been studied extensively \cite{?}. Denote by $\mathcal{Z}$ the set of all $z_\alpha(s,\xi)$.

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $w(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+3\rk\Phi$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $w(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $w(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
content...
\end{proof}
\noindent In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).

@@

We also define $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w(-1)$. Then
\[ h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta,\alpha\rangle}\xi\right),\quad \varepsilon\in R^*,\ \xi\in R. \]

@@

Denote by $\Hh(\Phi,R)$ the subgroup generated by all $h_\alpha(\varepsilon)$, $\alpha\in\Phi$, $\varepsilon\in R^*$. Then set
\[ \Hh(\Phi,R,I) = \Hh(\Phi,R)\cap\G(\Phi,R,I)=\langle h_\alpha(\varepsilon),\ \alpha\in\Phi,\ \varepsilon\in R^*\cap(1+I)\rangle. \]
\begin{lemma}
$\Hh(\Phi,R,I)\leqslant\E(\Phi,R,I)$.
\end{lemma}
\begin{proof}
Set $\varepsilon=1+s$, $s\in I$ and write
\begin{multline*}
h_\alpha(1+s) = x_\alpha\left(-1\middle)\, x_{-\alpha}\middle(-s\middle)\, x_\alpha\middle((1+s)^{-1}\middle)\, x_{-\alpha}\middle(s(1+s)\right) = \\
= x_\alpha\left((1+s)^{-1}-1\middle)\, z_{-\alpha}\middle(-s,(1+s)^{-1}\middle)\, x_{-\alpha}\middle(s(1+s)\right).
\end{multline*}
It remains to note that $(1+s)^{-1}\in 1+I$ and so all the factors lie in $\E(\Phi,R,I)$.
\end{proof}

>>>>

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system with a fixed basis of simple roots $\Pi$.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its {\it elementary subgroup}, 
 i.\,e. the subgroup generated by root unipotents $t_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the {\it relative Chevalley group} (also called {\it principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item {\it Gauss decomposition} i.\,e. decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item {\it Bass---Kolster} decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item {\it Dennis---Vaserstein} decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $t_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

%TODO: Illustrate each statement of the text below with a reference:
There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over small-dimensional rings.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, whereas Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The main advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions formulated in terms of the {\it stable rank} of the ground ring.
Their another advantage is that they prove useful in the study of asymptotical properties of Chevalley groups. We are going to discuss this theme in more detail below.

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the {\it width of $G$ with respect to $X$} i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $t_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
For example, W.~van der Kallen has shown that the width of $\SL_3(R)$ is infinite already for a euclidean ring $R=\mathbb{C}[t]$.

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
It is a classical result of J.~Tits and L.~Vaserstein that the relative elementary group $\E(\Phi, R, I)$ is generated by elements $z_\alpha(s,\xi)=t_{\alpha}(s)^{t_{-\alpha}(\xi)}$, see Theorem~\ref{theorem:Tits-Vaserstein} below.
Denote by $\mathcal{Z}$ the set of all $z_\alpha(s, \xi)$, $s\in I$, $\xi \in R$. 
Our first application of parabolic factorizations is the following estimate of the width of $\G(\Phi, R, I)$ in terms of $\mathcal{Z}$.

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $w(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $w(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $w(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_l$, $R = \mathbb{Z}$ of the above theorem was considered by U.~Hadad in~\cite[Theorem~1.6]{Ha12}.

@@

We also define $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w(-1)$. Then
\begin{equation}\label{rel:R6}
h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta,\alpha\rangle}\xi\right),\quad \varepsilon\in R^*,\ \xi\in R.
\end{equation}

@@

Denote by $\Hh(\Phi,R)$ the subgroup generated by all $h_\alpha(\varepsilon)$, $\alpha\in\Phi$, $\varepsilon\in R^*$. Then set
\[ \Hh(\Phi,R,I) = \Hh(\Phi,R)\cap\G(\Phi,R,I)=\langle h_\alpha(\varepsilon),\ \alpha\in\Phi,\ \varepsilon\in R^*\cap(1+I)\rangle. \]
\begin{lemma}\label{lemma:rel-tor-elementary}
$\Hh(\Phi,R,I)\leqslant\E(\Phi,R,I)$.
\end{lemma}
\begin{proof}
Set $\varepsilon=1+s$, $s\in I$ and write
\begin{multline*}
h_\alpha(1+s) = x_\alpha\left(-1\middle)\, x_{-\alpha}\middle(-s\middle)\, x_\alpha\middle((1+s)^{-1}\middle)\, x_{-\alpha}\middle(s(1+s)\right) = \\
= x_\alpha\left((1+s)^{-1}-1\middle)\, z_{-\alpha}\middle(-s,(1+s)^{-1}\middle)\, x_{-\alpha}\middle(s(1+s)\right).
\end{multline*}
It remains to note that $(1+s)^{-1}\in 1+I$ and so all the factors lie in $\E(\Phi,R,I)$.
\end{proof}
